---
title: "Pancreatic Cancer project"
author: "Isabella Oakes"
date: "6/15/2021"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r}
pancreatic <- read.csv('C:/Users/Ladybug/Desktop/pancreatic.csv', header = TRUE)
```

```{r}
summary(pancreatic)
```

```{r}
head(pancreatic)
```

```{r}
library(dplyr)
library(ggplot2)
pancreaticdata <- as.data.frame(pancreatic)
pancreaticdata$diagnosis <- as.factor(pancreaticdata$diagnosis)
levels(pancreaticdata$diagnosis) <- c("No pancreatic disease", "Benign hepatobiliary disease", "Pancreatic cancer")
qplot(patient_cohort, data = pancreaticdata, geom = "bar", fill = diagnosis) + 
  scale_fill_manual(values = c("#257272", "#efc234", "#af8db2"), name = "Diagnosis") +
  labs(title = "Patient Cohort Distribution", x = "Patient cohort", y = "Count") 
pancreaticdata %>% count(patient_cohort)
```


```{r}
qplot(sample_origin, data = pancreaticdata, geom = "bar", fill = diagnosis) + 
  scale_fill_manual(values = c("#257272", "#efc234", "#af8db2"), name = "Diagnosis") +
  labs(title = "Sample Origin Distribution", x = "Sample Origin", y = "Count") 
pancreaticdata %>% count(sample_origin)
```

```{r}
qplot(age, data = pancreaticdata, geom = "bar", fill = diagnosis) + 
  scale_fill_manual(values = c("#257272", "#efc234", "#af8db2"), name = "Diagnosis") +
  labs(title = "Age Distribution", x = "Age", y = "Count") 
```

```{r}
qplot(sex, data = pancreaticdata, geom = "bar", fill = diagnosis) + 
  scale_fill_manual(values = c("#257272", "#efc234", "#af8db2"), name = "Diagnosis") +
  labs(title = "Sex Distribution", x = "Sex", y = "Count") 
pancreaticdata %>% count(sex)
```

```{r}
qplot(diagnosis, data = pancreaticdata, geom = "bar", fill = diagnosis) + 
  scale_fill_manual(values = c("#257272", "#efc234", "#af8db2"), name = "Diagnosis") +
  labs(title = "Diagnosis Distribution", x = "Diagnosis", y = "Count") +
  theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1))
pancreaticdata %>% count(diagnosis)
```

```{r}
panstage <- pancreaticdata[pancreaticdata$diagnosis == "Pancreatic cancer", ]
qplot(stage, data = panstage, geom = "bar", fill = diagnosis) + 
  scale_fill_manual(values = c("#af8db2"), name = "Diagnosis") +
  labs(title = "Stage Distribution", x = "Stage", y = "Count") 
pancreaticdata %>% count(diagnosis)
```

```{r}
benign <- pancreaticdata[pancreaticdata$diagnosis == "Benign hepatobiliary disease", ]
qplot(benign_sample_diagnosis, data = benign, geom = "bar", fill = diagnosis) + 
  scale_fill_manual(values = c("#efc234"), name = "Diagnosis") +
  labs(title = "Benign Sample Diagnosis Distribution", x = "Benign Sample Diagnosis", y = "Count") 
benign %>% count(benign_sample_diagnosis)
```

```{r}
library(OneR)
plasma <- bin(pancreaticdata, nbins = 10, labels = NULL, method = c("length"), na.omit = TRUE)
qplot(plasma_CA19_9, data = plasma, geom = "bar", fill = diagnosis) + 
  scale_fill_manual(values = c("#257272", "#efc234", "#af8db2"), name = "Diagnosis") +
  labs(title = "Plasma CA19-9 Distribution", x = "Plasma CA19-9", y = "Count")
plasma %>% count(plasma_CA19_9)
```

```{r}
creatinine <- bin(pancreaticdata, nbins = 10, labels = NULL, method = c("length"), na.omit = TRUE)
qplot(creatinine, data = creatinine, geom = "bar", fill = diagnosis) + 
  scale_fill_manual(values = c("#257272", "#efc234", "#af8db2"), name = "Diagnosis") +
  labs(title = "Creatinine Distribution", x = "Creatinine", y = "Count") 
```

```{r}
LYVE1 <- bin(pancreaticdata, nbins = 15, labels = NULL, method = c("length"), na.omit = TRUE)
qplot(LYVE1, data = LYVE1, geom = "bar", fill = diagnosis) + 
  scale_fill_manual(values = c("#257272", "#efc234", "#af8db2"), name = "Diagnosis") +
  labs(title = "LYVE1 Distribution", x = "LYVE1", y = "Count") +
  theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1))
```

```{r}
REG1B <- bin(pancreaticdata, nbins = 15, labels = NULL, method = c("length"), na.omit = TRUE)
qplot(REG1B, data = REG1B, geom = "bar", fill = diagnosis) + 
  scale_fill_manual(values = c("#257272", "#efc234", "#af8db2"), name = "Diagnosis") +
  labs(title = "REG1B Distribution", x = "REG1B", y = "Count") +
  theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1))
```

```{r}
TFF1 <- bin(pancreaticdata, nbins = 15, labels = NULL, method = c("length"), na.omit = TRUE)
qplot(TFF1, data = TFF1, geom = "bar", fill = diagnosis) + 
  scale_fill_manual(values = c("#257272", "#efc234", "#af8db2"), name = "Diagnosis") +
  labs(title = "TFF1 Distribution", x = "TFF1", y = "Count") +
  theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1))
```


```{r}
REG1A <- bin(pancreaticdata, nbins = 15, labels = NULL, method = c("length"), na.omit = TRUE)
qplot(REG1A, data = REG1A, geom = "bar", fill = diagnosis) + 
  scale_fill_manual(values = c("#257272", "#efc234", "#af8db2"), name = "Diagnosis") +
  labs(title = "REG1A Distribution", x = "REG1A Origin", y = "Count") +
  theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1))
```

```{r}
library(corrplot)
pancreatic2 <- pancreatic
pancreatic2$sex <- as.factor(pancreatic2$sex)
pancreatic2$sex <- as.numeric(pancreatic2$sex)
corrpan <- pancreatic2[c(4:6, 9:14)]
corrpan[is.na(corrpan)] <- 0
cor(corrpan)

```


Data cleaning - missing values, degenerate distributions, highly correlated features


```{r}
library(DataExplorer)
#First handle missing values - first view the distribution of missing values
plot_missing(pancreatic)
```

Only missing values in two variables - plasma_CA19_9 and REG1A
REG1A - Urinary levels of a protein that may be associated with pancreas regeneration. Only assessed in 306 patients (one goal of the study was to assess REG1B vs REG1A) - this is why 48.14% of this variable is missing
plasma_CA19_9 - Blood plasma levels of CA 19–9 monoclonal antibody that is often elevated in patients with pancreatic cancer. Only assessed in 350 patients (one goal of the study was to compare various CA 19-9 cutpoints from a blood sample to the model developed using urinary samples). - again, missing because only assessed in a certain number of patients

```{r}
library(mice)
#Use MICE to impute missing values for both variables
tempDat <- mice(pancreatic, m = 5, maxit = 50, method = "pmm", seed = 27, printFlag = FALSE)
pancreatic_complete <- complete(tempDat, 2)
#Compare means of original variables and imputed variables using t-test
t.test(pancreatic$REG1A, pancreatic_complete$REG1A)
t.test(pancreatic$plasma_CA19_9, pancreatic_complete$plasma_CA19_9)
#Compare spread using standard deviation
c("Missing REG1A" = sd(pancreatic$REG1A, na.rm = TRUE), "No Missing REG1A" = sd(pancreatic_complete$REG1A))
c("Missing Plasma" = sd(pancreatic$plasma_CA19_9, na.rm = TRUE), "No Missing Plasma" = sd(pancreatic_complete$plasma_CA19_9))
```

Check if there are any missing values still
```{r}
#Check for missing values
sum(is.na(pancreatic_complete))
```

See if there are any degenerate features or highly correlated features
```{r}
library(caret)
#Check for near zero variance features
nearZeroVar(pancreatic_complete)
#Check for highly correlated features
findLinearCombos(pancreatic_complete[,c(4,9:14)])
```

Data preparation and splitting
```{r}
#Change the response variable diagnosis from three classes to two - pancreatic cancer (Yes) or no pancreatic cancer (No)
pancreatic_complete <- pancreatic_complete %>% 
  mutate(diagnosis = ifelse(diagnosis == 1 | diagnosis == 2, "No", "Yes"))

#Split the data into a train and a test set
samples <- createDataPartition(pancreatic_complete$diagnosis, p = 0.7, list = FALSE)[,1]
train <- pancreatic_complete[samples,]
test <- pancreatic_complete[-samples,]

#Ensure that the train and test have the same distribution for the response variable
sum(train$diagnosis == "Yes")/nrow(train)
sum(test$diagnosis == "Yes")/nrow(test)

#Set control parameters
ctrl <- trainControl(summaryFunction = twoClassSummary,
                     method = "cv")
```

```{r}
trainx <- train[c(-1, -6, -7, -8)]
trainy <- train[c(6)]
trainy <- as.matrix(trainy)
trainy <- as.factor(trainy)
ctrl2<- trainControl(summaryFunction = twoClassSummary,
                     method = "cv", classProbs = TRUE)
```

Random Forest Model

```{r}
set.seed(100)
RFmodel <- train(x = trainx, y = trainy,
                 method = "rf",
                 metric = "ROC",
                 trControl = ctrl2)
RFmodel
```

Random Forest Statistics 

```{r}
RFpred <- predict(RFmodel, newdata = test)
test$diagnosis <- as.factor(test$diagnosis)
confusionMatrix(RFpred, test$diagnosis)

```

```{r}
library(caret)
caret::varImp(RFmodel, scale = FALSE)
```

Naive Bayes model

```{r}
library(klaR)
library(MASS)
set.seed(100)
suppressWarnings({NBmodel <- train(x = trainx, y = trainy,
                 method = "nb",
                 metric = "ROC",
                 trControl = ctrl2)})
NBmodel
```

```{r}
suppressWarnings({NBpred <- predict(NBmodel, newdata = test)})
confusionMatrix(NBpred, test$diagnosis)

```
Nnet Model 1

```{r}
library(nnet)

set.seed(800)
nnetMod<- nnet(diagnosis~ age + creatinine + LYVE1, data = train, size = 3, decay = .1 )
nnetMod

pred<- predict(nnetMod, newdata= test, type="class")

cm1<- table(test$diagnosis, pred)
No_Accuracy<-(89/106)*100
No_Accuracy

confusionMatrix(cm1)

Model_Accuracy<- ((89+42)/(89+17+42+28))*100
Model_Accuracy
```

NNet Model

```{r}
library(nnet)

set.seed(800)


nnetgrid <- expand.grid(.size = 1:10,
                        .decay = c(0, .1, 1, 2))
maxsize <- max(nnetgrid$.size)
numwts <- 1*(maxsize * (length(train) +1) + maxsize + 1)
nnetmodel <- train(x = trainx, y = trainy,
                   method = "nnet",
                   metric = "ROC",
                   tuneGrid = nnetgrid,
                   trace = FALSE,
                   maxit = 200,
                   MaxNWts = numwts,
                   trControl = ctrl2)
nnetmodel

nnetpred<- predict(nnetmodel, newdata= test)
confusionMatrix(nnetpred, test$diagnosis)
```


```{r}
set.seed(800)
rpartfit<- train(x = trainx, y = trainy,
                 method = "rpart2",
                tuneLength = 10,
                trControl = ctrl2)
rpartfit
```

```{r}
test$diagnosis <- as.factor(test$diagnosis)
rpartpred <- predict(rpartfit, newdata = test)
confusionMatrix(rpartpred, test$diagnosis)
```

```{r}
predictions <- subset(test, select = c(6))
RFpreds <- predict(RFmodel, test, type = "prob")
predictions$RFprob <- RFpreds[,"No"]
predictions$RFclass <- predict(RFmodel, test)
```

```{r}
suppressWarnings({NBpreds <- predict(NBmodel, test, type = "prob")})
suppressWarnings({predictions$NBprob <- NBpreds[,"No"]})
suppressWarnings({predictions$NBclass <- predict(NBmodel, test)})
```

```{r}
nnetpreds <- predict(nnetmodel, test, type = "prob")
predictions$nnetprob <- nnetpreds[,"No"]
predictions$nnetclass <- predict(nnetmodel, test)
```

```{r}
rpartpreds <- predict(rpartfit, test, type = "prob")
predictions$rpartprob <- rpartpreds[,"No"]
predictions$rpartclass <- predict(rpartfit, test)
```

```{r}
predictions$diagnosis <- as.factor(predictions$diagnosis)
labs <- c(RFprob = "Random Forest",
          NBprob = "Naive Bayes",
          rpartprob = "Single Tree",
          nnetprob = "Neural Network")
liftCurve <- lift(diagnosis ~ RFprob + NBprob + rpartprob + nnetprob, data = predictions, labels = labs)
xyplot(liftCurve, auto.key = list(columns = 2, lines = TRUE, points = FALSE))
```









